---
title: "UNICEF Project Exploration - STA130 Winter 2024"
author: "Nikita Jain"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

# Final Project Overview: Identifying Opportunities to Accelerate Progress on Sustainable Development Goals (SDG)

## Guiding Research Question

How do landlocked countries compare with small island nations in their advancement toward meeting the UN’s Sustainable Development Goals (SDGs)?

## Specific Research Question

Do countries show higher/lower gender inequality due to their geographic categorization as landlocked or small islands based on the mean Women Empowerment Index scores?

## Data Wrangling and Cleaning

```{r}
# Feeding in file path
file_path <- "/home/rstudio/STA130-W24/Final Project/country_indicators.csv"

# load in country indicators with required variables
country_indicators <- 
  read_csv(file_path) %>%
  select(
    'iso3',
    'hdr_pr_f_2021',
    'sowc_maternal-and-newborn-health__demand-for-family-planning-satisfied-with-modern-methods-2016-2021-r_service-coverage-sub-index-on-reproductive-maternal-newborn-and-child-health',
    'sowc_adolescent-health__adolescent-birth-rate-2016-2021-r_aged-15-19_female',
    'sowc_women-s-economic-empowerment__educational-attainment-2008-2021-r_upper-secondary_female',
    'sowc_adolescents__transition-to-work-2013-2021-r_not-in-education-employment-or-training_female',
    'sowc_women-s-economic-empowerment__labour-force-participation-rate-2010-2020-r_female_total',
    'sowc_women-s-economic-empowerment__financial-inclusion-2014-2020-r_female_female',
    'sowc_adolescents__protection_intimate-partner-violence-2013-2020-r_female') %>% 
  select(iso3, everything()) %>%  
  rename(con_code = iso3) 

# preview data
country_indicators
```

sdg file

```{r}
# load SDG data and select necessary variable
sdg <- 
  read_csv("/home/rstudio/STA130-W24/Final Project/sdr_fd5e4b5a.csv") %>%
  select('Country Code ISO3') %>% rename(con_code = 'Country Code ISO3')

# preview data
sdg
```


country_codes file

```{r}
# load country_codes data and select variables
country_codes <- 
  read_csv("/home/rstudio/STA130-W24/Final Project/country_codes.csv") %>%
  select('ISO-alpha3 Code (M49)','Small Island Developing States (SIDS) (M49)', 
         'Land Locked Developing Countries (LLDC) (M49)') %>% 
  rename(con_code = 'ISO-alpha3 Code (M49)')

country_codes
```

## Data Integration

2- sample, 2-sided hypothesis testing

Null Hypothesis- There is no difference in the mean Women Empowerment Index scores between Landlocked and Small Islands
$H_0$: $g_1 = g_2$ and $g_1 - g_2 = 0$  

Alternate Hypothesis- There is a difference in the mean Women Empowerment Index scores between Landlocked and Small Islands
$H_1$: $g_1 != g_2$ and $g_1 - g_2 != 0$

The set significance level (alpha level) would be set to $\alpha$ = 0.05. Null hypothesis would be rejected if p <= $\alpha$

```{r}

# Integrating data from all three data sets using con_code as a common key

wei_rough_data <- inner_join(x=country_codes, y=country_indicators, by="con_code")
wei_clean <- inner_join(x=wei_rough_data, y=sdg, by="con_code")

# Renaming columns and removing extra variables

wei_rename <- wei_clean %>% 
  mutate(country_type = case_when(
      `Small Island Developing States (SIDS) (M49)` == TRUE ~ "Small Island",
      `Land Locked Developing Countries (LLDC) (M49)` == TRUE ~ "Land Locked",)) %>% 
  rename(
    MMC ='sowc_maternal-and-newborn-health__demand-for-family-planning-satisfied-with-modern-methods-2016-2021-r_service-coverage-sub-index-on-reproductive-maternal-newborn-and-child-health',
    ABR = 'sowc_adolescent-health__adolescent-birth-rate-2016-2021-r_aged-15-19_female',
    CSE = 'sowc_women-s-economic-empowerment__educational-attainment-2008-2021-r_upper-secondary_female',
    NEET = 'sowc_adolescents__transition-to-work-2013-2021-r_not-in-education-employment-or-training_female',
    LFPR = 'sowc_women-s-economic-empowerment__labour-force-participation-rate-2010-2020-r_female_total',
    FI = 'sowc_women-s-economic-empowerment__financial-inclusion-2014-2020-r_female_female',
    PR = hdr_pr_f_2021,
    IVP = 'sowc_adolescents__protection_intimate-partner-violence-2013-2020-r_female') %>% 
  select(con_code, country_type, everything(), -"Land Locked Developing Countries (LLDC) (M49)", 
         -"Small Island Developing States (SIDS) (M49)") 

# replacing missing values of FI and IVP with their mean scores
wei_rename_fill <- wei_rename %>% mutate(
    FI = if_else(is.na(FI), mean(FI, na.rm = TRUE), FI),
    IVP = if_else(is.na(IVP), mean(IVP, na.rm = TRUE), IVP)) %>% drop_na()


wei_rename_fill
```

# Creating Women Empowerment index-
```{r}

# Setting maximum and minimum values for each indicator taken from 
# Technical Note: Twin Indices on Women’s Empowerment and Gender Equality

min_MMC = 0
min_ABR = 0
min_CSE = 0
min_NEET = 0
min_LFPR = 0
min_FI = 0
min_PR = 0
min_IVP = 0

max_MMC = 100
max_ABR = 200
max_CSE = 100
max_NEET = 85
max_LFPR = 100
max_FI = 100
max_PR = 75
max_IVP = 60

# Normalizing each variable to help with comparison
# All variables are positive indicators (higher values indicate better performance in 
# that field) except ABR, NEET, IPV

wei_data <- wei_rename_fill %>% 
  mutate(
    
    # Normalizing positive indicators
    norm_MMC  = (MMC - min_MMC)/(max_MMC - min_MMC),
    norm_FI   = (FI - min_FI)/(max_FI - min_FI),
    norm_PR   = (PR - min_PR)/(max_PR - min_PR),
    norm_CSE  = (CSE - min_CSE)/(max_CSE - min_CSE),
    norm_LFPR = (LFPR - min_LFPR)/(max_LFPR - min_LFPR),
    
    # Normalizing negative indicators
    norm_ABR  = (max_ABR - ABR)/(max_ABR - min_ABR),
    norm_NEET = (max_NEET - NEET)/(max_NEET - min_NEET),
    norm_IVP  = (max_IVP - IVP)/(max_IVP - min_IVP))

# Calculation of dimension indices
wei_data <- wei_data %>% 
  mutate(
    I_health    = (norm_MMC + norm_ABR) / 2,
    I_education = (norm_CSE + norm_NEET) / 2,
    I_inclusion = (norm_LFPR + norm_FI) / 2,
    I_decision  = norm_PR,
    I_violence  = norm_IVP)
  

# Computing Women Empowerment Index (WEI is a positive index)
wei <- wei_data %>% select(-MMC, -ABR, -CSE, -NEET, -LFPR, -FI, -PR,-IVP) %>%
  mutate(WEI = (I_health * I_education * I_inclusion * I_decision * I_violence) ^ (1/5))

wei

```

# Calculating the observed test statistic- 
$\Delta\hat{g}$ will be the difference in the mean scores of WEI of landlocked and small island countries

```{r}
ghat <- wei %>% 
  group_by(country_type) %>%
  summarise(means = mean(WEI)) 
ghat

delta_ghat <- 
  wei %>% 
  group_by(country_type) %>%
  summarise(means = mean(WEI)) %>%
  summarise(value = diff(means)) %>%
  as.numeric()

print(delta_ghat)
```

\newpage
Below is R code that simulates $N = 1000$ values of the test statistic $\Delta\hat{g}_{\rm sim}$ **under the null hypothesis** using a permutation test. In this test, we assume that our groups are identical under our null hypothesis. Mixing the two groups together, randomly generating new groups with the same sizes, and then recomputing our test statistic each time therefore should allow us to simulate values from the sampling distribution provided our sample size is large enough.

```{r}
seed_num <- 130
set.seed(seed_num)  # creating seed

# setup
n_trials <- 1000  # number of permutations

# simulating test statistic (difference between mean scores of WEI)
delta_ghat_simulations <- rep(NA, n_trials)

for(i in 1:n_trials){
  # perform a random permutation
  simdata <- 
    wei %>%
    mutate(country_type = sample(country_type, replace=FALSE))
  
  # compute the simulated test statistic
  delta_ghat_sim <- 
    simdata %>% 
    group_by(country_type) %>%
    summarise(means = mean(WEI), .groups="drop") %>%
    summarise(value = diff(means)) %>%
    as.numeric()
  
  # store the simulated value
  delta_ghat_simulations[i] <- delta_ghat_sim
}

```

# Visualizations-

```{r}
# Visualizing sampling distribution of simulated test statistics using histograms

ggplot() +
  geom_histogram(aes(x=delta_ghat_simulations), color = "purple", fill = "purple", 
                 position = "Identity", alpha = 0.7, bins= 20) +
  labs(x = "Difference in Women Empowerment Index scores (delta_ghat_simulations)", 
       y = "Frequency",
       title = "Differences in 2-Sample Means for Gender Disparity Measurements") +
  theme_minimal() +
  theme(legend.position = "top")

# Creating box plots to compare the median WEI scores

ggplot(data = wei, aes(x = country_type, y = WEI, fill = country_type)) +
  geom_boxplot() +
  labs(x = "Country Type", y = "Women Empowerment Index Scores", 
       title = "Comparison of Women Empowerment Index Scores by Country Type") +
  scale_fill_manual(values = c("orange", "purple")) +  
  theme_minimal() +
  theme(legend.position = "top")

```


# Computing the p-value- 
(the probability of observing a test statistic at least as extreme as the observed value if the null hypothesis is true)
```{r}
# null hypothesis value
delta_median_null <- 0

p_value <-sum(abs(delta_ghat_simulations - delta_median_null) >= 
                abs(delta_ghat - delta_median_null)) / n_trials
print(p_value)

```

# Citations (MLA 9th edition)

1. Jain-Chandra, Sonali. “Chapter 2. Gender Inequality around the World.” Www.elibrary.imf.org, International Monetary Fund, 
www.elibrary.imf.org/display/book/9781513516103/ch002.xml#:~:text=The%20gender
%20gap%20varies%20strongly.  

2. TOWARDS IMPROVED MEASURES of GENDER INEQUALITY: An Evaluation of the UNDP Gender Inequality Index and a Proposal. 
www.unwomen.org/sites/default/files/2022-11/Discussion-paper-Towards-improved-measures-of-gender-inequality-en.pdf.   

3. Technical Note: Twin Indices on Women’s Empowerment and Gender Equality.
hdr.undp.org/sites/default/files/
publications/additional-files/2023-07/paths_equal_2023_tn.pdf.  
